Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-27 manuscript.tex
small changes and the PDF knitted

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Quantifying Stability and Change in Personal Culture Using Panel Data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\begin{document}
\maketitle
\begin{abstract}
Recent work has produced conflicting interpretations on the question of
whether adults undergo durable changes in their personal culture over
time. This paper asserts that this divergence is less due to
researchers' findings than the approach taken -- a tournament of models
focusing on the presence or absence of change. To move beyond this, we
reanalyze the data used previously in this debate with an approach that
quantifies the processes of cultural difference. In doing so, we compare
how much cultural difference is explained by intrapersonal change and
interpersonal differences. Our results harmonize recent findings by
showing that, while most measures of personal culture show change over
time, the variance explained by intrapersonal change is typically small
compared to interpersonal differences at baseline. Quantifying these
processes offers new ways to address theoretical questions of when and
how their relative importance shifts. We examplify this with the case of
college experience.
\end{abstract}

@@

A recurring focus in the sociology of culture is the question of whether
people, as they move through their lives, change their personal culture
--- the declarative and non-declarative attitudes, beliefs, values,
practices, and dispositions (Kiley and Vaisey 2020; Lersch 2023; Lizardo
2017).

This seemingly simple question underlies a number of central debates
across sociological subfields. For example, pragmatist theories of
action claim that changes in social environments cause people to adapt
their views and make new meanings (Gross 2009; Swidler 2001), while
Bourdieusian practice theories argue that the ``past conditions of
production'' leave a mark on people's personal culture that lasts
throughout their lives (Bourdieu 1990). Similarly, models of social
influence assume that people adapt their culture in the face of new
information (Vaisey and Lizardo 2010), while the emphasis on cohort
effects in models of aggregate social and cultural change requires them
to be open to change while young but become fairly resistant to it as
they age (Ryder 1965). Finally, life course theories posit slower but
still important changes over time as people advance through important
transitions in their lives (Bardi and Goodwin 2011; Elder and George
2016).

Because all of the above theoretical perspectives have some empirical
support, the current theoretical debate revolves around not whether they
exist, but when and how they each apply and the relative contribution of
each to explaining cultural differences. That being said, scholars have
struggled to reach a consensus on the relative importance of change
during adulthood. Over the span of a few years, most apparent changes in
people's responses to personal culture items appear to be transitory,
with little evidence of persistent change among adults (Kiley and Vaisey
2020; Vaisey and Kiley 2021). These results suggest change is not a
strong force in explaining personal cultural differences. When looked at
over longer time horizons, however, there is at least some evidence that
adults make persistent cultural changes, especially as they move through
distinct social roles across the life course (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating the different perspectives (Lersch 2023: p.~228).
Accordingly, researchers have formalized models of responses to cultural
items over time as demonstrating either ``change'' or ``no change'' to
then compare which model fits observed data better, thus functionally
asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only \emph{whether} people change
obscures what are actually highly compatible empirical patterns. Whether
researchers are able to detect ``change'' in a given dataset depends not
only on sample populations, item measurement, statistical methods and
power, but crucially also on definitions of what counts as ``change.''
These methodological differences therefore limit the possibility of
convergence on the theoretical debates. Moreover, asking whether these
is evidence of change limits the kind of questions that can be asked to,
``Do we see change in this measure?'' But such inquiries are ill-suited
to address what's at the core of the theoretical debate. There is enough
evidence to assume that neither socialization at an early age nor the
influence of the contemporary environment alone can explain cultural
differences. The question that remains is under what conditions, across
which groups, and in what domains do we observe differences in the
relative importance of intrapersonal change and interpersonal
differences. This focus is necessarily comparative and requires an
approach that quantifies forms of cultural change rather than calling
victory for the one or the other perspective.

In this paper, we try to move beyond previous debates by moving away
from the ``tournament of models'' approach toward a quantification of
the processes that lead to differences in personal culture. Drawing on
the seven panel datasets analyzed in these debates, we use established
methods to quantify the amount of variance in responses that is due to
between-people differences at a single point during the survey versus
the amount of variance in responses that is due to linear within-person
change over time. This measure gives us a sense of the importance of
intrapersonal change over the course of the panel relative to the
differences that exist between people assuming they never changed.

We observe similar patterns across the datasets that have produced
conflicting interpretations. In all datasets we see similar rates of
individual-level change over time across a broad range of measures of
personal culture. But quantifying these processes also reveals that
intrapersonal change plays a substantially smaller role in explaining
the variance in peoples' personal culture than between-person
differences in their baselines. Of course, whether this amount of
intrapersonal change is ``meaningful'' ultimately remains a theoretical
question reflected in prior expectations concerning the question,
population, and time. Nevertheless, with important exceptions, the
results suggest that intrapersonal change (in some cases over more than
10 years) does not seem to play a substantial role in explaining a broad
range of personal cultural differences we observe between adults.

These results help clarify and, hopefully, transcend previous debates.
What is more, the measure we propose offers a tool with wider utility
than merely discerning durable change that can help advance theoretical
debates. To exemplify this, we turn to the role of education for
cultural differences. Specifically, we use our measure to better
understand what mechanisms might underlie differences observed between
people with and those without college degrees. Focusing on eight
questions that tab general political dispositions, our measure enables
us to compare the relative importance of intrapersonal change for both
groups in ways that the tournament of models cannot. We find less
evidence of durable change among those with college degree. This may
imply that the college experience predominantly serves as a formative
period that solidifies political dispositions, rather than fostering an
openness {[}ADAPTABILITY?{]} to new information and influences later in
the life. {[}MISSING AN ENDING SENTENCE NOW.{]}

@@

Recent debates about whether adults undergo intrapersonal cultural
change emerged in part because theories of cultural change at the
aggregate level tend to implicitly invoke one of two models of
individual behavior. The first, what Kiley and Vaisey (2020) call a
``Settled Dispositions Model'' (SDM), assumes that peoples' personal
culture is relatively fixed by the time they are adults. While they
might make temporary changes in their declarative culture in reaction to
their environments, this model assumes that people return to a settled
baseline over a (relatively) short period of time. This model is
frequently invoked in theories of cultural change that suggest people
are strongly imprinted by early socialization experiences such as the
``past conditions of production'' a Bourdieusian practice theory
(Bourdieu 1990), cohort replacement theories of aggregate change
(Mannheim 1952; Ryder 1965), and control theories in social psychology
(Robinson 2007; Smith-Lovin and Heise 1988).

The second model summarized by Kiley and Vaisey (2020), an ``Active
Updating Model'' (AUM), posits that people continually update their
personal culture as they move through life. This model suggests people
change their personal culture as they adapt and make new meanings when
encountering new social environments, discourses, and information (Gross
2009; Swidler 2001). This model underlies, among others, theories of
cultural diffusion (Christakis and Fowler 2010), attitude alignment
(DellaPosta 2020), and polarization (Bail et al. 2018) It is also
implicit in most studies that ask whether specific experiences, changes
in social roles, or political events, affect personal culture (Gelman
and Margalit 2021; Slothuus and Bisgaard 2021; Visser and Mirabile
2004).

There is no reason to believe that only one of these two models is
``correct'' at all times and in all places. A population observed over
some period of time contains a mix of people who are noticeably changing
and people who are not. Instead, different perspectives argue that each
of these ideal-typical models is more operative at different times, for
different people, and for different elements of personal culture. For
example, adolescence and early adulthood is typically viewed as a
``formative period'' for personal cultural development and thus
characterized by higher rates of active updating, while middle age and
later life are potentially characterized more by settled dispositions
(Alwin and Krosnick 1991; Eaton et al. 2009; Krosnick and Alwin 1989).
Similarly, highly salient issues, such as views around gay rights in the
2010s; issues that see substantial elite realignments, such as views
around the Vietnam conflict in the 1970s (Zaller 1992); and novel issues
of public opinion, such as views around vaccines during the Covid-19
pandemic, might be characterized by active updating, while established
issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the AUM
was preferred for the majority of items, the total amount of durable
change detected on these items was small. A substantial minority of
questions (39 percent) favored the SDM, meaning they showed no evidence
of durable change. Questions with more evidence of durable change
included highly salient issues like gay marriage and ``public'' issues
like partisan identification and religious service attendance. There was
also more evidence of durable change among early adults (people ages
18-30) than the rest of the population. Overall, the researchers
concluded that ``results ultimately suggest that real, persistent
attitude change is an uncommon phenomenon among adults'\,' (Kiley and
Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This lack of
durable change is consistent with other findings that cohort replacement
plays a larger role than period effects in explaining differences in
personal culture (Vaisey and Lizardo 2016).

@@

Drawing on these findings, Lersch challenged the SDM and AUM as a
``needless dichotomy'' (2023). Instead he proposed the ``Life Course
Adaption Model'' (LCAM). This model draws on the life course perspective
to model personal culture as a linear trajectory over the course of a
panel that varies for each respondent. Lersch rectifies two shortcomings
of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey,
posits that changes follow a Markovian process where personal culture
responses at time \(t\) rely solely on responses at time \(t-1\).
Howevre, earlier life experiences can (directly and indirectly) mold
personal culture when transitioning to new social roles or into new
environments, even if their initial impact is delayed. For example,
childhood events might influence views on family structures later when
individuals form their own families. The LCAM considers therefore
considers influences from earlier than just \(t-1\) on responses at time
\(t\).

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on
three-wave panel data over four years, which might not be extensive
enough to properly adjudicate {[}COMPARE? WE ARE USING ADJUDICATING A
LOT{]} the two models. Lersch, however, evaluates the LCAM against the
AUM and SDM using panel data spanning a wider duration (from 3 to 36
years) and more waves (3 to 18), offering a better change to observe
durable change.

In Lersch's tournament of models, the LCAM gains broad {[}EXTENSIVE?{]}
support from data across five countries. Out of 428 questions, the LCAM
was favored in 297 cases while the SDM was preferred in only 112. The
rest did not yield inconclusive results, and notably, no questions
favoured the AUM. He concludes that ``new experiences over the life
course {[}\ldots{]} can persistently move individuals' personal culture
in novel directions'' (Lersch 2023: p.~243-244).

Despite differing interpretations of these scholars, their empirical
results align closely. Kiley and Vaisey (2020) discovered that most
questions they investigated favored the AUM, indicating evidence of
durable change among adults -- a finding largely in line with Lersch's.
However, while Lersch (2023) found a preference for the LCAM in many
cases, durable change only accounted for a minor proportion of variance
explained. On average, individuals changed {[}CAN WE SAY SHIFTED TO
INDICATE THE LINEAR DRIFT{]} about .07 standard deviations over 10
years, echoing Kiley's and Vaisey's view that such changes are typically
modest. Furthermore, a quarter of the items in Lersch's study favoured
the SDM, suggesting that despite broader assumptions about change and
more extensive data, many questions reveal no evidence of durable
change. Likewise, in studies claiming evidence of durable change in
personal culture, the magnitude is often minimal. For instance, Brocic
and Miles (2021) note that obtaining graduate degrees in humanities,
arts, and social sciences only shift {[}NUDGES?{]} peoples' moral
relativism by about .2 standard deviations compared to non-degree
holders {[}DELETED THE LAST SENTENCE PART HERE; FELT UNNECESSARY{]}.
Moreover, studies of aggregate change demonstrate that even on questions
where cohort effects prevail, there is always evidence that some
individuals change over time (Vaisey and Lizardo 2016). In other words,
despite different interpretations, the results of previous work are in
many ways highly consonant.

@@

Current debates often narrow down to ask whether people \emph{ever}
change, which is rarely theoretically interesting since the answer is
almost always ``yes.'' Over sufficient time and in a large enough
sample, researchers will likely observe some evidence for durable
changes in individuals. A lack thereof may simply due to a lack of
clarity in the question asked or a low level of resolution in the
possible answers, or that the question simply was not asked long enough
or to enough people. Conversely, finding evidence of change tells us
little about how much change is present in a population and how
meaningful that change is for social differences.

A more theoretically productive approach begins with a model assuming
that during an observed time period some individuals might remain
stable, might change a little, or might undergo significant shifts in
their personal culture. Lersch's LCAM formally does this by modeling
each individual as following their own trajectory over time. But rather
than debating whether this model is superior to a model that assumes
that people never change, researchers should quantify individual changes
over time and compare those to the differences that exist between people
at the panel's start. In other words, the focus should shift from
asking, ``do people change?'' to asking, ``how do change and initial
differences collectively explain variation in personal culture?'' {[}I
CHANGED THAT QUESTION WORDING, PLEASE CHECK!{]}

In this approach, two metrics reflect distinct theoretical processes.
First, ``early imprinting,'' as Lersch calls it, denotes interpersonal
differences at the start of the panel, capturing the accumulated
experiences of people before participating in the panel that influence
their answers. While commonly associated with experiences during an
early formative period that result in settled disposition, this could
also reflect transformative experiences (2014) that happen at any stage
in life, as long as they predate the panel and consistently impact
subsequent responses. For instance, for those people entering the panel
post-retirement, this ``imprinting'' might reflect this pivotal life
transition. Consequently, this metric also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Quantifying Stability and Change in Personal Culture Using Panel Data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\begin{document}
\maketitle
\begin{abstract}
Recent work has produced conflicting interpretations on the question of
whether adults undergo durable changes in their personal culture over
time. This paper asserts that this divergence is less due to
researchers' findings than the approach taken -- a tournament of models
focusing on the presence or absence of change. To move beyond this, we
reanalyze the data used previously in this debate with an approach that
quantifies the processes of cultural difference. In doing so, we compare
how much cultural difference is explained by intrapersonal change and
interpersonal differences. Our results harmonize recent findings by
showing that, while most measures of personal culture show change over
time, the variance explained by intrapersonal change is typically small
compared to interpersonal differences at baseline. Quantifying these
processes offers new ways to address theoretical questions of when and
how their relative importance shifts. We examplify this approach by
exploring how the importance of these processes differs between college
graduates and the rest of the population.
\end{abstract}

@@

A recurring focus in the sociology of culture is the question of whether
people, as they move through their lives and transition between social
roles, change their personal culture --- the declarative and
non-declarative attitudes, beliefs, values, practices, and dispositions
that exist at the individual level (Kiley and Vaisey 2020; Lersch 2023;
Lizardo 2017).

This seemingly simple question underlies a number of debates across
sociological subfields. For example, pragmatist theories of action claim
that changes in social environments cause people to adapt their views
and make new cultural meanings (Gross 2009; Swidler 2001), while
Bourdieusian practice theories argue that the ``past conditions of
production'' leave a mark on people's personal culture that lasts
throughout their lives (Bourdieu 1990). Similarly, models of social
influence assume that people adapt their culture in the face of new
information (Vaisey and Lizardo 2010 KK Note: I don't feel like this is
a good cite for this.), while the emphasis on cohort effects in models
of aggregate social and cultural change requires them to be open to
change while young but become fairly resistant to it as they age (Ryder
1965). Finally, life course theories posit slower but still important
changes over time as people advance through important transitions in
their lives (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Because all of the above theoretical perspectives have some empirical
support, the current debate revolves around not whether they exist, but
when and how they each apply and the contribution of each to explaining
cultural differences. That being said, scholars have struggled to reach
a consensus on the relative importance of change during adulthood. Over
the span of a few years, most apparent changes in people's responses to
personal culture items appear to be transitory, with little evidence of
persistent change among adults (Kiley and Vaisey 2020; Vaisey and Kiley
2021). This suggests that change during adulthood is not a strong force
in explaining personal cultural differences. When looked at over longer
time horizons, however, there is at least some evidence that adults make
persistent cultural changes, especially as they move through distinct
social roles (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating different theoretical perspectives (Lersch 2023: p.~228).
Accordingly, researchers have formalized models of responses to cultural
items over time as demonstrating either ``change'' or ``no change,''
which essentially asks whether we \emph{ever} observe intrapersonal
change in personal culture.

We argue that asking only \emph{whether} people change obscures what are
actually highly compatible empirical patterns. Being able to detect
change in a given dataset depends not only on sample populations, item
measurement, statistical methods, and power, but crucially also on
definitions of what counts as ``change.'' These methodological decisions
potentially limit the possibility of convergence on the theoretical
debates. Moreover, asking whether there is evidence of change limits the
kind of questions that can be asked to, ``Do we see change in this
measure?'' But such inquiries are ill-suited to address what's at the
core of the theoretical debate. There is enough evidence to assume that
neither socialization at an early age nor the influence of the
contemporary environment alone can explain cultural differences. The
question that remains is under what conditions, across which groups, and
in what domains do we observe differences in the relative importance of
intrapersonal change and interpersonal differences. This focus is
necessarily comparative and requires an approach that quantifies forms
of cultural change rather than calling victory for the one or the other
perspective.

In this paper, we try to intervene in this debate by moving beyond the
``tournament of models'' approach toward a quantification of the
processes that lead to differences in personal culture. Drawing on the
seven panel datasets analyzed in the debate, we use established methods
to quantify the amount of variance in responses that is due to
between-people differences at a single point during the survey versus
the amount of variance in responses that is due to linear intrapersonal
change over time. This measure gives us a sense of the importance of
change over the duration of the panel relative to the differences that
exist between people assuming they never changed.

We observe similar patterns across the datasets that yielded previously
conflicting interpretations. In all datasets we find a similar
distribution of how much explanatory power can be attributed to
intrapersonal change. But quantifying these processes also reveals that
intrapersonal change plays a substantially smaller role in explaining
the variance in peoples' personal culture than interpersonal differences
at their baselines. Whether the amount of change is large enough to be
``meaningful'' ultimately remains a theoretical question reflected in
prior expectations about particular cultural item and population. It
also crucially depends on how long people are observed in a panel.
Nevertheless, on many measures of personal culture we do not observe
enough intrapersonal change, in some cases over more than 10 years, for
it to play a substantial role in explaining the differences we observe
between adults, though notable exceptions exist.

These results help clarify and, hopefully, transcend previous debates
about whether adults change. What is more, the approach we propose can
be applied to a broader range of theoretical questions than approaches
discerning whether durable change exists in a panel. To exemplify this,
we explore differences in the explanatory power of intrapersonal change
for people with college degrees and those without. We find that change
has less explanatory power for people with college degrees than those
without. This may imply that the college experience solidifies political
dispositions rather than fosters an openness and adaptability to new
information and influences later in life. This offers a novel, empirical
basis to theorize about the role of college in shaping personal culture.
We believe that this approach has the potential to advance the
theoretical debate on the role of environments, social identities, and
roles in facilitating personal cultural change and stability.

@@

Recent debates about whether adults undergo intrapersonal cultural
change emerged in part because theories of cultural change at the
aggregate level tend to implicitly invoke one of two models of
individual behavior. The first, what Kiley and Vaisey (2020) call a
``Settled Dispositions Model'' (SDM), assumes that peoples' personal
culture is relatively fixed by the time they are adults. While they
might make temporary changes in their declarative culture in reaction to
their environments, this model assumes that people return to a settled
baseline over a short period of time. This model underlies theories of
cultural change that suggest people are imprinted by early socialization
experiences such as the ``past conditions of production'' a Bourdieusian
practice theory (Bourdieu 1990), cohort replacement theories of
aggregate change (Mannheim 1952; Ryder 1965), and control theories in
social psychology (Robinson 2007; Smith-Lovin and Heise 1988).

The second model summarized by Kiley and Vaisey (2020), an ``Active
Updating Model'' (AUM), posits that people continually update their
personal culture as they move through life. This model suggests people
change their personal culture as they adapt and make new meanings when
encountering new social environments, discourses, and information (Gross
2009; Swidler 2001). This model underlies, among others, theories of
cultural diffusion (Christakis and Fowler 2010), attitude alignment
(DellaPosta 2020), and polarization (Bail et al. 2018). It is also
implicit in most studies that ask whether specific experiences, changes
in social roles, or political events, affect personal culture (Gelman
and Margalit 2021; Slothuus and Bisgaard 2021; Visser and Mirabile
2004).

There is no reason to believe that only one of these two models is
``correct'' at all times and in all places. A population observed over
some period of time contains a mix of people who are noticeably changing
and people who are not. Instead, different perspectives argue that each
of these ideal-typical models is more operative at different times, for
different people, and for different elements of personal culture. For
example, adolescence and early adulthood is typically viewed as a
``formative period'' for personal cultural development and thus
characterized by higher rates of active updating, while middle age and
later life are potentially characterized more by settled dispositions
(Alwin and Krosnick 1991; Eaton et al. 2009; Krosnick and Alwin 1989).
Similarly, salient issues, such as views around gay rights in the 2010s;
issues that see substantial elite realignments, such as views around the
Vietnam conflict in the 1970s (Zaller 1992); and novel issues of public
opinion, such as views around vaccines during the Covid-19 pandemic
(Scoville et al. 2022), might be characterized by active updating, while
established issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items preferred the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they showed no evidence of durable
change. Questions with more evidence of durable change included salient
issues like gay marriage and questions tapping ``public'' statements or
behaviors such as partisan identification and religious service
attendance. There was also more evidence of durable change among early
adults (people ages 18-30) than the rest of the population. Overall, the
researchers concluded that ``results ultimately suggest that real,
persistent attitude change is an uncommon phenomenon among adults'\,'
(Kiley and Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This
lack of durable change is consistent with other recent findings that
cohort replacement plays a larger role than period effects in explaining
differences in personal culture (Vaisey and Lizardo 2016).

@@

Drawing on these findings, Lersch (2023) challenged the SDM and AUM as a
``needless dichotomy,'' proposing the ``Life Course Adaption Model''
(LCAM) as an alternative. This model draws on the life course
perspective to model personal culture as a different linear trajectory
over the duration of a panel for each respondent. In doing this, Lersch
rectified two shortcomings of the AUM and SDM. First, the AUM, as
described by Kiley and Vaisey, posits that changes follow a Markov
process where responses at time \(t\) are a function of responses at
time \(t-1\) but not earlier time points. However, earlier life
experiences can (directly and indirectly) mold personal culture when
transitioning to new social roles or into new environments, even if
their initial impact is delayed. For example, childhood events might
influence views on family structures later when individuals form their
own families. Therefore, the LCAM considers influences from earlier than
just \(t-1\) on responses at time \(t\).

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on
three-wave panel data over four years, which might not be extensive
enough to adjudicate the two models. Lersch evaluates the LCAM against
the AUM and SDM using panel data spanning a wider duration (from 3 to 36
years) and more waves (3 to 18), offering a better chance to observe
durable change.

When Lersch compared the LCAM to the AUM and SDM on data from five
countries, 297 of the 428 questions he analyzed preferred the LCAM,
suggesting that we observe some linear change for adults on most items.
The SDM was preferred on 112 items, and the rest did not yield
conclusive results. No questions favored the AUM. He concludes that
``new experiences over the life course {[}\ldots{]} can persistently
move individuals' personal culture in novel directions'' (Lersch 2023:
p.~243-244).

Despite differences in how these researchers interpret their findings,
the empirical results are not far apart. Kiley and Vaisey (2020) found
that the majority of items they tested favored the AUM, meaning there
was evidence of durable change among adults, a pattern consistent with
Lersch's results. While Lersch (2023) found evidence of durable change
on most items, this change only accounted for a small proportion of
variance explained. On average, people changed only about .07 standard
deviations over 10 years. This is consistent with Kiley and Vaisey's
finding that durable change is often small. And a quarter of the items
Lersch studied still preferred the SDM, suggesting that even with
different assumptions assumptions about change and more extensive data,
many questions show no evidence of durable change. Likewise, in studies
showing evidence of durable change in specific measures of personal
culture, the change is often small. For example, Brocic and Miles (2021)
estimate that completing graduate degrees in humanities, arts, and
social sciences shift peoples' moral relativism only about .2 standard
deviations on average compared to people with no college degree, and
this is the largest effect they identify. And studies of aggregate
change show that, even on items where cohort effects explain more
variance than period effects, there is always evidence that some people
change over time (Vaisey and Lizardo 2016). In other words, despite
different interpretations, the results of previous work are in many ways
highly consonant. {[}I think the revisions that were done to this
paragraph were confusing and mostly reverted it.{]}

@@

Current debates often narrow down to ask whether people \emph{ever}
change, which is rarely theoretically interesting since the answer is
almost always ``yes.'' Over sufficient time and in a large enough
sample, researchers will likely observe some evidence for durable
changes in individuals. A lack of evidence for change may simply due to
a lack of clarity in the question asked or a low level of resolution in
the possible answers, or that the question simply was not asked for long
enough or to enough people. Conversely, finding evidence of change tells
us little about how much change is present in a population and how
meaningful that change is for social differences.

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM formally does this by modeling each individual as
following their own trajectory over time. But rather than debating
whether this model fits a particular data set better than a model that
assumes that people never change, researchers should quantify individual
changes over time and compare those to the differences that exist
between people at the panel's start. In other words, the focus should
shift from asking, ``do people change?'' to asking, ``what are the
relative contributions of change and initial differences toward
explaining differences in personal culture?'' {[}KK Note: I didn't like
this change. It's not about collectively explaining, its about them
relative to each other.{]}
